---
title: "Workshop 8 TEMPLATE"
subtitle: "9 am"
author: "Stephen"
date: "2026-05-2026"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show upl
---

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?  

**Specific leaf area** is a continuous variable measured in cm^2^/g.  

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).  

**Species** is a categorical variable (again, a *factor*) with 6 levels.  

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| _Encelia californica_        | ENCCAL       | Bush sunflower     |  
| _Eschsholzia californica_    | ESCCAL       | California poppy   | 
| _Penstemon centranthifolius_ | PENCEN       | Scarlet bugler     | 
| _Grindelia camporums_        | GRINCAM      | Gumweed            | 
| _Salvia leucophylla_         | SALLEU       | Purple sage        | 
| _Stipa pulchra_              | STIPUL       | Purple needlegrass | 
| _Lotus scoparius_            | LOTSCO       | Deerweed           |

# 1. Set up

```{r}
#| label: packages and data

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here

drought_exp <- read_xlsx(
  here("data", 
       "Valliere_etal_EcoApps_Data.xlsx"),
  sheet = "First Harvest"
)

```

```{r}
#| label: plot colors and aesthetics

# storing colors to use for species
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

plant_colors <- c(
  "Lotus scoparius" = lotsco_col,
  "Penstemon centranthifolius" = pencen_col,
  "Salvia leucophylla" = salleu_col,
  "Encelia californica" = enccal_col,
  "Stipa pulchra" = stipul_col,
  "Eschsholzia californica" = esccal_col,
  "Grindelia camporums" = gricam_col)

# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_bw())
```


# 2. Clean data

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)

```{r}
#| label: data cleaning

# make object drought_exp_clean
drought_exp_clean <- drought_exp |> 
  # clean names: makes lowercase and underscores columns
  clean_names() |>
  # Rename scientific names under column species_name
  mutate(species_name = recode_values(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
  )) |>
  # Creating column with full name for treatment
  mutate(water_treatment = case_match(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"
  ))  |>
  # making order of species levels
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |>
  # reorder water treatment levels
  mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |>
  # keep only species name, water treatment, sla , and total g column
  select(species_name, water_treatment, sla, total_g)
```

Insert code to double check the structure of the data frame here.  
```{r}
str(drought_exp_clean)
```


Insert code to display 10 random rows from the data frame here.  


```{r}
#| label: 10 Rows

slice_sample(
  drought_exp_clean, # data frame to display
  n = 10 # selecting 10 rows from data frame
)
```



# 3. Visualizing data

## What is the relationship between SLA and plant mass?  

Make a scatterplot to visualize SLA (the predictor) on the x-axis, and total mass (the response) on the y-axis.  

Label the axes with units where appropriate.  

Add a title describing the relationship (e.g. as SLA increases, total mass ______). 

```{r}
#| label: Scatterplot

# base layer GG Plot
ggplot(data = drought_exp_clean,
       # x-axis: specific Leaf Area
       # y-axis: total plant mass
       mapping = aes(x = sla,
                     y = total_g))+
  # second layer: geom point
  geom_point()+
  # label y and x-axes and adding title
  labs(x = "Specific Leaf Area (cm\U00B2 /g)",
       y = "Total Mass g",
       title = "As SLA Increases Total Mass (g) Increases" )
```


## What are the differences in total mass between water treatments?

Make a jitterplot to visualize the differences in plant mass between water treatments.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between water treatments.

```{r}
#| label: Jitterplot

# base layer: gg ploy
ggplot(data = drought_exp_clean,
       mapping = aes(x = water_treatment , # x-axis: treatment
                     y = total_g,# y-axis: plant mass
                     color = water_treatment)) + #color by water treatment
  #second layer jitter plot                
  geom_jitter(width = 0.1,
              height = 0) +
  # set colors by water treatment
  scale_color_manual(
    values = c("Drought stressed" = ds_col,
               "Well watered" = ww_col)
  ) +
  # label axes and title
  labs(x = "Treatments",
       y = "Total Mass (g)",
       title = "On average, well watered plants have higher mass\nthan drought stressed plants") +
  # remove legend
  theme(legend.position = "none")
```





## What are the differences in total mass between species?

Make a jitterplot to visualize the differences in plant mass between species.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between species.  

Bonus: order the x-axis by decreasing mean plant mass.

```{r}
#| label: Jitter-Species

ggplot(data = drought_exp_clean,
       mapping = aes(x = species_name,
                     y = total_g,
                     color = species_name)) +
  geom_jitter(width = 0.1,
              height = 0) +
  scale_color_manual(values = plant_colors) +
  scale_x_discrete(labels = label_wrap(10)) +
  labs(x = "Species name",
       y = "Total plant mass (g)",
       title = "Species Differ in Total Plant Mass") +
  theme(legend.position = "none")
```


# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species | Predictor list              |  
|--------------|:---:|:---------------:|:-------:|:---------------------------:|  
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |       X         |    X    | all predictors (full model) | 
| 2            |  X  |       X         |         | SLA and water treatment     |   
| 3            |  X  |                 |    X    | SLA and species             |    
| 4            |     |       X         |    X    | water treatment and species |    
| 5            |  X  |                 |         | SLA                         |    
| 6            |     |       X         |         | water treatment             |    
| 7            |     |                 |    X    | species                     |

## Model fitting

```{r}
#| label: model fitting

# model 0: null model
model0 <- lm(
  total_g ~ 1, #formula, no predictors
  data = drought_exp_clean # clean data frame
)

# model 1: all predictors
model1 <- lm(
  total_g ~ sla + water_treatment + species_name, #formula, no predictors
  data = drought_exp_clean # clean data frame
)

# model 2: SLA and water treatment
model2 <- lm(
  total_g ~ sla + water_treatment,
  data = drought_exp_clean
)

# model 3: SLA and species
model3 <- lm(
  total_g ~ sla + species_name,
  data = drought_exp_clean
)

# model 4: water treatment and species
model4 <- lm(
  total_g ~ species_name + water_treatment,
  data = drought_exp_clean
)

# model 5: SLA
model5 <- lm(
  total_g ~ sla,
  data = drought_exp_clean
)

# model 6: water treatment
model6 <- lm(
  total_g ~ water_treatment,
  data = drought_exp_clean
)

# model 7: species
model7 <- lm(
  total_g ~ species_name,
  data = drought_exp_clean
)

```

## Model diagnostics

```{r}
#| label: Check Diagnostics

par(mfrow = c(2,2))
# generating diagnostic plots
    plot(model4)
```

-residuals appear normally distributed
-residuals appear homoscedastic (have constant varince)
-no clear outlier in data set influencing coefficient estimate



## Model selection

Insert code to use the `AICc()` function to calculate the AIC for each model.  

Arrange by decreasing AIC.

```{r}
#| label: calculating-aic

# calculating AIC
AICc(
  model0,
  model1,
  model2,
  model3,
  model4,
  model5,
  model6,
  model7
) |>
  # arrange output in order of AIC
  arrange(AICc)
```



Insert code to display the summary for the best model. 





## Stop and think: what does this model mean?  

**What is the best model?**

The best model includes water treatment and species as predictors

**How much variation in the response (total mass, in grams) does this model explain?**

72.6% of the variation in plant mass is explained by a model with both of these predictors

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

-Treatment influences plant mass
-Species influences plant mass

# 4. Model predictions

Insert code to generate model predictions for the best model.  

```{r}
#| Label: Summary Model 4
summary(model4)
```
-"(intercept)": model prediction for plant mass for Lotus scoparius in the drought stressed treatment

- "water_treatmentWell watered": model prediction for the difference in Lotus scoparius mass between the well watered and drought stressed treatment.

-"species_namePenstemon centranthifolius": model prediction for the difference between Pensteman mass in the well watered treatment and Lotus mass in the well watered treatment

-"species_nameEschscholzia californica": model predicition for the difference between Eschscholzia mass in the well watered treatment and Lotus mass in the well watered treatment.

Rename the `x` column and `group` column to match the original data frame (`drought_exp_clean`).

Look at the object before plotting.

```{r}
preds <- ggpredict(
  model4,
  terms = c("water_treatment", "species_name")
) |>
  rename(water_treatment = x,
species_name = y)
```


# 5. Final figure

Create a clean, publication ready figure depicting predictions for the best model and the underlying data.

```{r}
ggplot(data = drought_exp_clean,
       mapping = aes(x = water_treatment,
                     y = total_g,
                     color = water_treatment))+
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21,
              alpha = 0.8) +
  geom_pointrange(data = preds,
                  mapping = aes(x = water_treatment,
                                y = predicted,
                                ymin = conf.low,
                                ymax = conf.high)) +
  facet_wrap(~ species_name)
```

# 6. Writing

## Figure caption

## Basic components of test

**predictors**  


**response**  


**test**  


**distribution**  


**parameters (degrees of freedom)**  


**R^2^**  


**p-value**   


**significance level**  


**reference level(s)**  


## Written communication

### Methods

We used linear models to determine the predictive relationship between specific leaf area (cm^2^/g), treatment, and species on plant mass (g). We evaluated model assumptions of normally distributed errors using QQ plots of model residuals. We evaluated assumptions of constant variance of errors using Residuals vs Fitted and Scale-Location plots. We used Akaike's Information Criterion (AIC) to select the simplest model that explained the most variation in plant mass.  

### Results

We found that the simplest model describing the greatest variation in plant mass included treatment and species as predictors (linear model: F(2, 67) = 14.57, p < 0.001, $\alpha$ = 0.05). 







